---
title: "Estimating extinction dates from sighting records"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating extinction dates from sighting records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

## Abstract

When a species has not been reliably sighted for a long time, two
questions follow: is it still extant, and if not, when did it most likely
disappear. Both questions can be addressed from the sighting record alone,
without additional biological data, if a null model of the sighting
process is assumed. This vignette summarizes the six estimators
implemented in EDE, applies them to a single worked example, and discusses
where and why they disagree.

## The problem

A sighting record is a sequence of confirmed observations of a species
over time, typically sparse and irregular. Extinction is rarely directly
observed; what is observed is the absence of further sightings after some
last confirmed date. Two related problems arise from this kind of record:

1. **Persistence testing.** Given a candidate year, is the absence of
   sightings since the last confirmed one statistically surprising under
   the assumption that the species is still present and being observed at
   its historical rate? If so, persistence is rejected.
2. **Point and interval estimation.** Rather than testing a single
   candidate year, estimate the most likely extinction date directly, with
   a measure of uncertainty when possible.

The estimators below differ in which of these two problems they solve,
and in what they assume about the process that generates sightings.

## The estimators

**Robson & Whitlock (1964).** Extrapolates from the gap between the two
most recent sightings, scaled by `(1 - alpha) / alpha`. Assumes the
sighting process behaves, near the true endpoint, like a uniform record
process. No confidence interval; a single point estimate.

**Strauss & Sadler (1989).** A classical confidence bound on the true
endpoint of a temporal range, given only the first and last sighting and
the number of sightings in between, under a uniform-occurrence model
within the range. Was developed for stratigraphic ranges and applies
directly to sighting ranges. Fast, but uses only three summary numbers
from the record and ignores its internal spacing.

**Roberts & Solow (2003), `ole()`.** The best linear unbiased estimator
under a Weibull-type record-value model fit to the spacing of the largest
sighting times. Uses every recorded sighting time, not just the extremes,
and returns both a point estimate and a confidence interval. Requires at
least three sightings with a positive count.

**Solow (1993).** A nonparametric persistence test. Under a homogeneous
Poisson sighting process, the ratio of "time since last sighting" to
"total observation window" has a known null distribution; raising it to
the power of the total sighting count gives the chance of persistence at
a candidate year. Simple, and does not require knowing when in the window
each sighting occurred beyond the last one, but treats sighting effort as
constant over time.

**Solow (2005).** The same logic as Solow (1993), but the observation
window is weighted by cumulative sighting effort (sighting count times
time since the first sighting) instead of raw elapsed time. This matters
when sighting effort was not constant: a record with many early sightings
and very few late ones is not equally informative per year throughout.

**Burgman, Grimson & Ferson (1995).** A combinatorial persistence test.
If sightings were placed uniformly at random across the candidate window,
what is the probability that the largest observed gap between sightings
would not exceed the gap actually seen. Computed via an
inclusion-exclusion argument over Stirling numbers of the second kind.
Makes no distributional assumption beyond uniform placement, at the cost
of being the most computationally expensive of the six.

`solow1993()`, `solow2005()`, and `burgman1995()` all return either the
first candidate year at which the chance of persistence drops to or below
`alpha`, or, with `data_out = TRUE`, the full curve of chance against
candidate year.

## A worked example

The dataset below is illustrative, built to resemble a typical
sighting record for a rare, possibly extinct species: frequent early
sightings, increasingly sparse later ones, and a long silence after the
last confirmed observation.

```{r}
library(EDE)

years <- c(1900, 1902, 1903, 1905, 1907, 1908, 1910, 1912, 1915, 1918,
           1920, 1923, 1925, 1928, 1930, 1933, 1936)
sightings <- c(4, 3, 5, 2, 3, 4, 2, 1, 2, 1, 1, 2, 1, 1, 1, 1, 1)

sd <- sighting_data(data.frame(year = years, sightings = sightings))
sd
```

The last confirmed sighting is 1936. Sighting frequency was already
declining well before that: the median gap between sightings is 2 years
up to 1915, and grows to 3 years afterward.

### Running all six estimators

```{r}
ole(sd)
robson1964(sd)
strauss1989(sd)
solow1993(sd, test_year = 2000)
solow2005(sd, test_year = 2000)
burgman1995(sd, test_year = 2000)
```

| Method | Estimate | Notes |
|---|---|---|
| Robson & Whitlock (1964) | 1993 | extrapolates only from the last two sightings (1933, 1936); with a 3-year gap and alpha = 0.05, extrapolates 57 years forward |
| Strauss & Sadler (1989) | 1943.4 | uses only the range (1900-1936) and the sighting count |
| Roberts & Solow (2003) | 1940.6, CI [1936.3, 1954.0] | uses the full spacing of all 17 sighting times |
| Solow (1993) | 1940 | first year the persistence chance drops below 0.05, weighting the window by elapsed time |
| Solow (2005) | not rejected by 2000 | same test, weighted by sighting effort instead |
| Burgman, Grimson & Ferson (1995) | 1944 | first year the gap-occupancy chance drops below 0.05, on a decreasing run |

### Why Solow (1993) and Solow (2005) disagree

```{r}
curve_1993 <- solow1993(sd, test_year = 2000, data_out = TRUE)
curve_2005 <- solow2005(sd, test_year = 2000, data_out = TRUE)

plot(curve_1993$time, curve_1993$chance, type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, 1), xlab = "candidate extinction year", ylab = "chance of persistence")
lines(curve_2005$time, curve_2005$chance, lwd = 2, col = "firebrick")
abline(h = 0.05, lty = 2, col = "grey50")
legend("topright", legend = c("Solow (1993)", "Solow (2005)", "alpha = 0.05"),
       col = c("steelblue", "firebrick", "grey50"), lty = c(1, 1, 2), lwd = c(2, 2, 1), bty = "n")
```

Solow (1993) treats every year of the observation window as equally
informative, so the chance of persistence decays steadily as the silence
after 1936 lengthens, crossing 0.05 in 1940. Solow (2005) instead weights
each year by how much sighting effort it represents, inferred from the
sighting record itself. Because sightings were already sparse well before
1936, the model attributes less evidential weight to the recent silence,
and the chance of persistence flattens out around 0.21 instead of
continuing to zero. Neither curve is wrong; they encode different
assumptions about whether the observation process was stationary. A
record with a long history of *consistently frequent* sightings up to the
last one would show the two curves converge, since sighting effort would
look roughly constant throughout.

### Why Robson & Whitlock (1964) gives such a late estimate here

Robson & Whitlock only look at the two most recent sighting times, 1933
and 1936. A 3-year gap scaled by `(1 - 0.05) / 0.05 = 19` places the
estimate at `1936 + 3 * 19 = 1993`. This estimator is sensitive to the
specific spacing of the last two observations, and can be pulled far in
either direction by a single unusually short or long final gap. It is
best read alongside an estimator like `ole()` that draws on the entire
sighting history rather than just the last interval.

## Choosing among the estimators

There is no single correct estimator for every record. As a starting
point:

- If the record is long and reasonably well sampled throughout, `ole()`
  gives a point estimate with an interval and uses the most information.
- If sighting effort clearly declined over time, prefer `solow2005()`
  over `solow1993()`; both answer "is persistence still plausible", not
  "when did extinction occur".
- If only the endpoints and total count are available, `strauss1989()` is
  the minimal-assumption option.
- `robson1964()` is cheap to compute but rests entirely on the last
  interval; treat it as a sanity check, not a primary estimate.
- `burgman1995()` makes the weakest distributional assumption but is the
  slowest, since it evaluates an inclusion-exclusion sum for every
  candidate year.

In practice, running more than one estimator and checking whether they
agree is informative in itself: sharp disagreement, as in the Solow
comparison above, usually points to a specific feature of the record
(here, declining sighting effort) rather than to a flaw in either method.

## References

Burgman, M. A., Grimson, R. C., & Ferson, S. (1995). Inferring threat from
scientific collections. *Conservation Biology*, 9(4), 923-928.

Roberts, D. L., & Solow, A. R. (2003). Flightless birds: When did the
dodo become extinct? *Nature*, 426(6964), 245.

Robson, D. S., & Whitlock, J. H. (1964). Estimation of a truncation
point. *Biometrika*, 51(1/2), 33-39.

Solow, A. R. (1993). Inferring extinction from sighting data. *Ecology*,
74(3), 962-964.

Solow, A. R. (2005). Inferring extinction from a sighting record.
*Mathematical Biosciences*, 195(1), 47-55.

Strauss, D., & Sadler, P. M. (1989). Classical confidence intervals and
Bayesian probability estimates for ends of local taxon ranges.
*Mathematical Geology*, 21(4), 411-427.
